
%-------------------------------------
% Resum in LaTeX (XeLaTeX)
% Author:  HF Yuan
% Project: https://github.com/xyz-yuanhf/yuan-resume
% Base on: https://github.com/mattyHerzig/mattys_resume
%          http://www.jianxu.net/en/
% License: MIT
%------------------------------------

\documentclass[a4paper, 10pt]{article}
\usepackage{myresume}  % Style package
\usepackage{hyperref}

% --------------------  START  --------------------
\begin{document}

% -------------------- HEADING --------------------
\begin{flushright}
  \setstretch{0.6}
  \item {\Calluna aaron[dot]graybill[at]stanford.edu}  % E-mail address
  \item {\Calluna \url{https://aarongraybill.com}}  % Home page
\end{flushright}\vspace{-45pt}

\begin{flushleft}  % Rplace here with your name (and identity if required)
  {\Atkinson \fontsize{30pt}{30pt}\selectfont \textsc{Aaron Graybill}} \quad {\Atkinson \fontsize{14.5pt}{14.5pt}\selectfont \textsc{}}
  \noindent\rule{\textwidth}{0.4pt}
\end{flushleft}

% -------------------- EDUCATION --------------------
\sectionBlock{
\section{Education}
}{
\eduHeading
    {Stanford University Graduate School of Business}{Stanford, California, USA}
    {PhD Student, Quantitative Marketing}{2025--Present}
% \itemListStart
%     \namedItem{Relevant Coursework: }{Microeconomics}
% \itemListEnd

\eduHeading
  {Haverford College}{Haverford, PA, USA}
  {B.A. in Economics and Mathematics}{2018--2022}
\itemListStart
  \namedItem{Relevant Economics Coursework: }{Advanced Microeconomics, Mathematical Economics, Quantitative Economics, Thesis}
  \namedItem{Relevant Math Coursework: }{Analysis, Linear Algebra, Stochastic and Numerical Methods, Quantitative Finance, Probability, Applied Multivariate Statistical Analysis}
\itemListEnd

}

% -------------------- PUBLICATIONS --------------------
% \sectionBlock{
% \section{Publications}
% }{
% \pubListStart
% \justifying
% \item \textbf{Xiao Yuan}, Hua Li. The Future Urban Transportation Systems: Innovations and Challenges. \textit{Journal of Operations Research and Optimization}, 2024.
% \item Hua Li, \textbf{Xiao Yuan}, John Doe. Optimizing Logistics and Supply Chain Networks Using Machine Learning Techniques. \textit{International Conference on Operations Research and Machine Learning}, 2023.
% \item John Doe, \textbf{Xiao Yuan}, Hua Li. Artificial Intelligence in Healthcare: Transforming Diagnostics and Treatment. \textit{International Conference on HealthTech Innovations}, 2023.
% \pubListEnd
% }

% -------------------- PROJECTS --------------------
\sectionBlock{
\section{Professional\\Experience}
}{
\profeshHeading
{Research Analyst, Federal Reserve Bank of Philadelphia}{Philadelphia, PA, USA}{Jun. 2022--Present}
\itemListStart
\myItem{
    Developed novel methodology to uniquely identify homeowners between properties using co-ownership record and mailing address history
    }
\myItem{
    Leveraged concepts from network theory and linear algebra to convert methodology into optimized algorithms
}
\myItem{
    Authored more than 5,000 lines of R and SQL code to apply algorithms to hundreds of gigabytes of property-level tax assessment and transaction data
}
\myItem{
    Created new spatial data to measure how fragmentation of parcel use types affects long-term housing development
}
\myItem{
    Automated chart production in Python for Monetary Policy briefings
    }
\myItem{
    Modernized internal data collection and processing by scraping available data, removing reliance on 3rd party vendors
    }
\itemListEnd

\profeshHeading
{Volunteer Alumnus Teaching Assistant, Haverford College}{Haverford, PA, USA}{Jan. 2023--May 2023}
\itemListStart
\myItem{
    Lead problem solving sessions for Advanced Microeconomics, the terminal microeconomics course
}
\myItem{
    Adapted to student needs, teaching concepts from probability theory, calculus, and intermediate microeconomics
}
\itemListEnd

\profeshHeading
{Research Intern, Federal Reserve Bank of Philadelphia}{Philadelphia, PA, USA}{Jun. 2020--May 2022}
\itemListStart
\myItem{
    Analyzed survey results from the Manufacturing Business Outlook Survey for the publication of monthly reports about the local business climate
}
\myItem{
    Generated a dashboard of statistics and charts on COVID-19’s economic impacts
}
\itemListEnd



\profeshHeading
{Teaching Assistant, Haverford College}{Haverford, PA, USA}{Sep. 2021--Dec. 2021}
\itemListStart
\myItem{
    Answered student questions about econometrics
}
\myItem{
    Graded problem sets and provided constructive commentary
}
\itemListEnd

}

% -------------------- INTERNSHIPS --------------------
% \sectionBlock{
% \section{Internships}
% }{
% \internHeading
%   {ABC Tech Ltd.}{Shanghai, China}{2024.01 - 2024.06}
% \itemListStart
%   \myItem{Develop engaging content for social media platforms.}
%   \myItem{Prepare reports and presentations summarizing research findings.}
% \itemListEnd
% \internHeading
%   {XYZ Tech Inc.}{Shanghai, China}{2023.07 - 2023.12}
% \itemListStart
%   \myItem{Develop engaging content for social media platforms.}
%   \myItem{Prepare reports and presentations summarizing research findings.}
% \itemListEnd
% }

% -------------------- AWARDS & HONORS --------------------
\sectionBlock{
\section{Awards\\and\\Honors}
}{
\awardListStart
\justifying
\item \emph{Michael M. Weinstein Prize in Economics}, Haverford College \hfill May 2022\\
For having made the greatest overall contribution to departmental life. Awarded for contributions in class, with peers, and for starting the widely used Economics Question Center.
\item \emph{Runner-Up, Holland Hunter Thesis Prize}, Haverford College \hfill May 2022\\
For outstanding senior theses in economics. One of three selected as finalists for the award.
\awardListEnd
}

% -------------------- SKILLS --------------------
\sectionBlock{
\section{Skills}
}{
\skillListStart
\justifying
\item \emph{Programming}: R, Python, SQL, Terminal (Bash,  PowerShell), Julia, Stata, Matlab.
\item \emph{Technical}: git (GitHub/GitLab), Apache Hadoop, \TeX.
\skillListEnd
}

% -------------------- OTHERS --------------------
% \sectionBlock{
% \section{Academic Services}
% }{
% \skillListStart
% \justifying
% \item \emph{Reviewers for}: \textit{Journal of Operations Research and Optimization},
% \item \phantom{\emph{Reviewers for}:} \textit{International Conference on Optimization and Machine Learning},
% \item \phantom{\emph{Reviewers for}:} \;...
% \skillListEnd
% }

\end{document}

